### 2020-07-17 paper.tex
Notation (replace A with K)

\textbf{The space of finite dictionaries.}
We fix a countably infinite set $\mathcal{A}$ of possible keys, such that each key $a$ is either called \emph{discrete} ($a \in \mathcal{I}$) or \emph{continuous} ($a \in \mathcal{J}$), where $\mathcal{A} = \mathcal{I} \cup \mathcal{J}$.\footnote{
It is possible to assign a general measure space to each address, but this is not necessary for our purposes.}
Let $V_a$ denote the set of possible values for key $a$, where $V_a$ is a countable set for each discrete key, and where $V_a = \mathbb{R}^{d_a}$ for each continuous key for some $d_a$.
Given a set of keys $A$, let $V_A = \times_{a \in A} V_a$ denote the set of \textit{assignments} of values to each key.
Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in V_A\}$. That is, a dictionary specifies 
a finite set of keys $A \subset \mathcal{A}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_a$ for each.

@@

\textbf{A measure space of finite dictionaries.}
We associate a measure $\mu_a$ on $V_a$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_a}$ for each continuous key.
For each finite set of keys $A$, we make $V_A$ a measure space using the standard product $\sigma$-algebra $\Sigma_A = \otimes_{a \in A} \Sigma_a$ and the product measure $\mu_A = \times_{a \in A} \mu_a$.
We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{A \subset \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) \mid \mathbf{x} \in B_A\} \mid B_A \in \Sigma_A \, \text{for each finite } A \subset \mathcal{A}\}$ to obtain a measurable space
of dictionaries.
A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_A$: we set $\mu(B) := \sum_{A \in \mathcal{A}, |A| < \infty} \mu_A(\{\mathbf{x} \mid (A, \mathbf{x}) \in B\})$.

\textbf{Notation for dictionaries.} Given a dictionary $d = (A, \mathbf{x})$, we write $A_d$ for $A$ and $d[a]$ for the value $x_a$ associated with a key $a \in A$.
We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
For example, the dictionary $(A, \mathbf{x})$ with $A = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.

\textbf{Probability distributions on finite dictionaries.} 
When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
The probability is distributed the finite sets of keys $A \subseteq \mathcal{A}$ where $|A| < \infty$:
\begin{equation}
1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{A \subseteq \mathcal{A}\\|A| < \infty}} \int p(m) \mu_{A}(dm)
\end{equation}
where the probability for each key set $A$ is:
\begin{equation}\label{eq:integral}
\int p(m) \mu_{A}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((A, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
\end{equation}
where $d: = \sum_{a \in A \cap \mathcal{J}} d_{a}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $A$ and $\mathbf{x}_2$ is an assignment to the continuous choices.

We now give an example to build intuition.
Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$. If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density on dictionaries $d$ is:
%
\begin{align*}
p(d) =
\begin{array}{l}
p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
\end{array}
\end{align*}
%
when $A_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other addresses $a \in \mathcal{A}$, $V_a = \mathbb{R}$ with the Lebesgue measure for $\mu_a$.
For each $k \in \{0, 1, \ldots\}$, the probability for key set
$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$
is $p_{\mathrm{poisson}(s)}(k)$ (via Equation~(\ref{eq:integral})).

\textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
%For a set of keys $A \subset \mathcal{A}$, let $p(A) := \int 1[A_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $A$. 
We say a key $a \in \mathcal{A}$ \textit{almost always appears} if $\int \mathbf{1}[a \in A_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$. Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$. Furthermore, let $(A_m, \mathbf{m}) \oplus (A_n, \mathbf{n}) := (A_m \cup A_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $A_m$ and $A_n$. Then we can define the conditional density $p(d \mid b) := \mathbf{1}[A_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid A_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite. If the reference measure $\mu_a$ associated with each key $a \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all addresses in $B$). When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{chang1997conditioning}.

Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}_P$ and $\mathcal{A}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{A}_P$ and similarly for $\mathcal{D}_Q$.
% TODO: this is only needed for Algorithm 2, and it is relevant for the 
To simplify the notation, and without loss of generality, we will assume that $\mathcal{A}_P$ and $\mathcal{A}_Q$ are disjoint\footnote{If $\mathcal{A}_P$ and $\mathcal{A}_Q$ are not disjoint then, they can be made so by adding a different prefix to each address.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{A}_P \cup \mathcal{A}_Q$
(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).

Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant.
Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation*}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation*}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
\[
z_1 \sim z_2 \iff (A_{z_1} = A_{z_2}) \land (z_1[a] = z_2[a] \; \forall a \in A_{z_1} \cap \mathcal{I})
\]
(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

\begin{algorithm*}[h]
\begin{algorithmic}
\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$
\vspace{1mm}
\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$; compute log-density via probabilistic program}
    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
    \State $\log p(x) \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density via probabilistic program}
    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density via probabilistic program}
    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
    \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$  \algorithmicend
\EndProcedure\\
\vspace{1mm}
\begin{minipage}[t]{0.49\linewidth}
\Procedure{trace-and-score}{$\mathcal{P}$}
   \State $s \gets 0$
   \State $x \gets \{\}$
   \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
        \State 1. set $v \sim \mathrm{distribution}$
        \State 2. set $x[a] \gets v$
        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
        \State 4. evaluate to $v$)
    \END
   \State \Return $(x, s)$
\EndProcedure
\end{minipage}%
\hfill
\begin{minipage}[t]{0.49\linewidth}
\Procedure{score}{$\mathcal{P}$, $x$}
    \State $s \gets 0$
    \State $A \gets\{\}$
    \execute[$\mathcal{P}$], but with $\mathtt{"}a \sim \mathrm{distribution}\mathtt{"} \equiv$ (
        \State 1. set $v \gets x[a]$
        \State 2. set $A \gets A \cup \{a\}$
        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
        \State 4. evaluate to $v$)
    \END
   \State \algorithmicif\hspace{1mm} $A \ne A_x$ \algorithmicthen\hspace{1mm} $s \gets -\infty$ \algorithmicend
   \State \Return $s$
\EndProcedure
\end{minipage}\\
\vspace{1mm}
\Procedure{run-involution}{$\mathcal{F}$, $z$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..
    \LineComment{Run the involution, and keep track of copied, read, and written continuous keys}
    \State $z' \gets \{\}$ \Comment{Initialize empty output dictionary}
    \execute[$\mathcal{F}$], but with
        \State $\mathtt{"write}(a, v)\mathtt{"} \equiv$
            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
            \algorithmicthen\hspace{1mm} set $z'[a] \gets v$
            \algorithmicelse\hspace{1mm} set $z'[a] \gets v$ and set $W \gets W \cup \{a\})$
        \State $\mathtt{"read}(a)\mathtt{"} \equiv$
            (\algorithmicif\hspace{1mm} $a \in \mathcal{I}$
            \algorithmicthen\hspace{1mm} evaluate to $z[a]$
            \algorithmicelse\hspace{1mm} set $R \gets R \cup \{a\}$ and evaluate to $z[a]$)
        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (set $z'[a_2] \gets z[a_1]$ and set $C \gets C \cup \{a_1\}$)
    \END
    \LineComment{Use automatic differentiation of $\mathcal{F}$ to compute Jacobian, skipping copied addresses}
    \For{$i$ in $1$ to $|W|$}
        \State $a_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ w.r.t. non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
    \EndFor
    \State \Return $(z', \log(|\det(J)|))$
\EndProcedure
\end{algorithmic}
\caption{Automated Involutive MCMC}
\label{alg:auto-involutive-mcmc}
\end{algorithm*}

@@

At the most basic level, a probabilistic program is a program that makes random choices. 
Any such program induces a probability distribution over its possible \textit{execution traces}, records of each random choice it makes.
If each random choice is associated with a unique \textit{address} from a countable set
of addresses $\mathcal{A}$, 
then these traces can be viewed as finite dictionaries, mapping the address of each random choice to its value. 
The distribution induced by a probabilistic program over
its execution traces can thus be understood as a measure on the space $(\mathcal{D}, \Sigma_\mathcal{D}, \mu_\mathcal{D})$ introduced in the previous section. Furthermore, densities of trace distributions with respect to $\mu_\mathcal{D}$ are typically easy to compute.

@@

%the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\

@@

%then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\

@@

%\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}

>>>>

\textbf{The space of finite dictionaries.}
We fix a countably infinite set $\mathcal{K}$ of possible keys, such that each key $k$ is either called \emph{discrete} ($k \in \mathcal{I}$) or \emph{continuous} ($k \in \mathcal{J}$), where $\mathcal{K} = \mathcal{I} \cup \mathcal{J}$.\footnote{
It is possible to assign a general measure space to each key, but this is not necessary for our purposes.}
Let $V_k$ denote the set of possible values for key $k$, where $V_k$ is a countable set for each discrete key, and where $V_k = \mathbb{R}^{d_k}$ for each continuous key for some $d_k$.
Given a set of keys $K$, let $V_K = \times_{k \in A} V_k$ denote the set of \textit{assignments} of values to each key.
Then the set of all finite dictionaries is $\mathcal{D} := \bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in V_K\}$. That is, a dictionary specifies 
a finite set of keys $K \subset \mathcal{K}$ at which it has values, and an assignment $\mathbf{x}$ of values $x_k$ for each.

@@

\textbf{A measure space of finite dictionaries.}
We associate a measure $\mu_k$ on $V_k$ for each key---the counting measure for each discrete key and the Lebesgue-measure on $\mathbb{R}^{d_k}$ for each continuous key.
For each finite set of keys $K$, we make $V_K$ a measure space using the standard product $\sigma$-algebra $\Sigma_K = \otimes_{k \in A} \Sigma_k$ and the product measure $\mu_K = \times_{k \in A} \mu_k$.
We equip $\mathcal{D}$ with the $\sigma$-algebra $\Sigma_{\mathcal{D}} := \{\bigcup_{K \subset \mathcal{K}, |K| < \infty} \{(K, \mathbf{x}) \mid \mathbf{x} \in B_K\} \mid B_K \in \Sigma_K \, \text{for each finite } K \subset \mathcal{K}\}$ to obtain a measurable space
of dictionaries.
A reference measure $\mu_\mathcal{D}$ on this space can be constructed using the product measures $\mu_K$: we set $\mu(B) := \sum_{A \in \mathcal{K}, |K| < \infty} \mu_K(\{\mathbf{x} \mid (K, \mathbf{x}) \in B\})$.

\textbf{Notation for dictionaries.} Given a dictionary $d = (K, \mathbf{x})$, we write $K_d$ for $K$ and $d[k]$ for the value $x_k$ associated with a key $k \in K$.
We also denote specific dictionaries using notation $\{k_1 \mapsto v_1, k_2 \mapsto v_2, \ldots\}$.
For example, the dictionary $(K, \mathbf{x})$ with $K = \{1, \mathtt{"foo"}\}$ and $x_1 = 0.123$ and $x_{\mathtt{"foo"}} = 5$ is denoted 
$\{1 \mapsto 0.123, \mathtt{"foo"} \mapsto 5\}$.

\textbf{Probability distributions on finite dictionaries.} 
When all keys are discrete, a probability distribution on dictionaries is defined by a $p$ is a probability mass function $p : \mathcal{D} \to [0, 1]$ that assigns a probability $p(m)$ to each dictionary $m \in \mathcal{P}$ such that $\sum_{x \in \mathcal{D}} p(m) = 1$.
More generally a probability distribution on dictionaries is defined by a probability density $p : \mathcal{D} \rightarrow [0, \infty)$ such that $\int p(m) \mu_\mathcal{D}(\text{d}m) = 1$. 
The probability is distributed the finite sets of keys $K \subseteq \mathcal{K}$ where $|K| < \infty$:
\begin{equation}
1 = \int p(m) \mu_\mathcal{D}(\text{d}m) = \sum_{\substack{K \subseteq \mathcal{K}\\|K| < \infty}} \int p(m) \mu_{K}(dm)
\end{equation}
where the probability for each key set $K$ is:
\begin{equation}\label{eq:integral}
\int p(m) \mu_{K}(dm) = \sum_{\mathbf{x}_1} \int_{\mathbb{R}^{d}} p((K, (\mathbf{x}_1, \mathbf{x}_2))) d\mathbf{x}_2
\end{equation}
where $d: = \sum_{k \in A \cap \mathcal{J}} d_{k}$ and where $\mathbf{x}_1$ is an assignment to the discrete choices in $K$ and $\mathbf{x}_2$ is an assignment to the continuous choices.

We now give an example to build intuition.
Consider a generative model of univariate data points $y_1, \dots, y_n$ from a Gaussian mixture with an unknown number of components $k$, each with unknown mean $m_i$ and variance $s_i$. If we place a Gaussian prior on $m_i$, an inverse Gamma prior on $s_i$, and a Poisson prior on $k$, the resulting density on dictionaries $d$ is:
%
\begin{align*}
p(d) =
\begin{array}{l}
p_{\mathrm{poisson}(3)}(d[\mathtt{k}]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{normal}(0, 1)}(d[\mathtt{m}_i]) \cdot\\
\prod_{i=1}^{d[\mathtt{k}]} p_{\mathrm{inversegamma}(1, 10)}(d[\mathtt{s}_i]) \cdot\\
\prod_{i=1}^n  \frac{1}{d[\mathtt{k}]} \sum_{j=1}^{d[\mathtt{k}]} \cdot p_{\mathrm{normal}(d[\mathtt{m}_j], d[\mathtt{s}_i])}(d[\mathtt{y}_i])
\end{array}
\end{align*}
%
when $K_d = \{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq d[\texttt{k}]\} \cup \{\texttt{s}_i \mid 1 \leq i \leq d[\texttt{k}]\}$, and 0 otherwise. In this case, we have $V_\texttt{k} = \mathbb{N}$ with the counting measure for $\mu_\texttt{k}$, and for all other keys $k \in \mathcal{K}$, $V_k = \mathbb{R}$ with the Lebesgue measure for $\mu_k$.
For each $k \in \{0, 1, \ldots\}$, the probability for key set
$\{\texttt{k}, \texttt{y}_\texttt{1}, \dots, \texttt{y}_\texttt{n}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$
is $p_{\mathrm{poisson}(s)}(k)$ (via Equation~(\ref{eq:integral})).

\textbf{Conditional distributions via disintegration.} Consider a probability density $p$ on the space $\mathcal{D}$ of dictionaries. 
%For a set of keys $K \subset \mathcal{K}$, let $p(A) := \int 1[K_m = A] p(m) \mu_\mathcal{D}(\text{d}m)$, the probability under $p$ of a dictionary with exactly the key set $K$. 
We say a key $k \in \mathcal{K}$ \textit{almost always appears} if $\int \mathbf{1}[k \in K_m] p(m) \mu_\mathcal{D}(\text{d}m) = 1$. Suppose $B$ is a set of keys that almost always appear for $p$, and that $b = (B, \mathbf{b})$ is a dictionary with keys $B$. Furthermore, let $(K_m, \mathbf{m}) \oplus (K_n, \mathbf{n}) := (K_m \cup K_n, \mathbf{m} \times \mathbf{n})$ be the \textit{merge} of two dictionaries $m$ and $n$ defined on disjoint key sets $K_m$ and $K_n$. Then we can define the conditional density $p(d \mid b) := \mathbf{1}[K_d \cap B = \emptyset] \frac{p(d \oplus b)}{\int_{\{m \mid K_m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)}$ when the denominator is finite. If the reference measure $\mu_k$ associated with each key $k \in B$ is discrete (e.g., the counting measure), then this definition corresponds to the ordinary notion of conditioning on an event (namely, the event that a sample from $p$ agrees with the dictionary $b$ on all keys in $B$). When this is not the case, it corresponds to a more general measure-theoretic notion called disintegration~\citep{chang1997conditioning}.

Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
The conditional density $p(d | b)$ is nonzero only if $K_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{K}_P$ and $\mathcal{K}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{K}_P$ and similarly for $\mathcal{D}_Q$.
% TODO: this is only needed for Algorithm 2, and it is relevant for the 
To simplify the notation, and without loss of generality, we will assume that $\mathcal{K}_P$ and $\mathcal{K}_Q$ are disjoint\footnote{If $\mathcal{K}_P$ and $\mathcal{K}_Q$ are not disjoint then, they can be made so by adding a different prefix to the keys of each set.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{K}_P \cup \mathcal{K}_Q$
(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).

Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $K_x$ and $K_y$ and the values $x_k$ and $y_k$ for all discrete keys $k$ are both constant.
Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation*}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation*}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
\[
z_1 \sim z_2 \iff (K_{z_1} = K_{z_2}) \land (z_1[k] = z_2[k] \; \forall k \in K_{z_1} \cap \mathcal{I})
\]
(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

\begin{algorithm*}[h]
\begin{algorithmic}
\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$
\vspace{1mm}
\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$; compute log-density via probabilistic program}
    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
    \State $\log p(x) \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density via probabilistic program}
    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density via probabilistic program}
    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
    \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$  \algorithmicend
\EndProcedure\\
\vspace{1mm}
\begin{minipage}[t]{0.49\linewidth}
\Procedure{trace-and-score}{$\mathcal{P}$}
   \State $s \gets 0$
   \State $x \gets \{\}$
   \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
        \State 1. set $v \sim \mathrm{distribution}$
        \State 2. set $x[k] \gets v$
        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
        \State 4. evaluate to $v$)
    \END
   \State \Return $(x, s)$
\EndProcedure
\end{minipage}%
\hfill
\begin{minipage}[t]{0.49\linewidth}
\Procedure{score}{$\mathcal{P}$, $x$}
    \State $s \gets 0$
    \State $K \gets\{\}$
    \execute[$\mathcal{P}$], but with $\mathtt{"}k \sim \mathrm{distribution}\mathtt{"} \equiv$ (
        \State 1. set $v \gets x[k]$
        \State 2. set $K \gets K \cup \{k\}$
        \State 3. set $s \gets s + \textproc{logpdf}(\mathrm{distribution}, v)$
        \State 4. evaluate to $v$)
    \END
   \State \algorithmicif\hspace{1mm} $K \ne K_x$ \algorithmicthen\hspace{1mm} $s \gets -\infty$ \algorithmicend
   \State \Return $s$
\EndProcedure
\end{minipage}\\
\vspace{1mm}
\Procedure{run-involution}{$\mathcal{F}$, $z$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..
    \LineComment{Run the involution, and keep track of copied, read, and written continuous keys}
    \State $z' \gets \{\}$ \Comment{Initialize empty output dictionary}
    \execute[$\mathcal{F}$], but with
        \State $\mathtt{"write}(k, v)\mathtt{"} \equiv$
            (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
            \algorithmicthen\hspace{1mm} set $z'[k] \gets v$
            \algorithmicelse\hspace{1mm} set $z'[k] \gets v$ and set $W \gets W \cup \{k\})$
        \State $\mathtt{"read}(k)\mathtt{"} \equiv$
            (\algorithmicif\hspace{1mm} $k \in \mathcal{I}$
            \algorithmicthen\hspace{1mm} evaluate to $z[k]$
            \algorithmicelse\hspace{1mm} set $R \gets R \cup \{k\}$ and evaluate to $z[k]$)
        \State $\mathtt{"copy}(k_1, k_2)\mathtt{"} \equiv$ (set $z'[k_2] \gets z[k_1]$ and set $C \gets C \cup \{k_1\}$)
    \END
    \LineComment{Use automatic differentiation of $\mathcal{F}$ to compute Jacobian, skipping copied addresses}
    \For{$i$ in $1$ to $|W|$}
        \State $k_i \gets W_i$ \Comment{Pick $i$th key in $W$; the order does not matter}
        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[k_i] = h_{e(z)}[k_i]$ w.r.t. non-copied continuous inputs ($z_k$ for $k \in R \setminus C$)}}
    \EndFor
    \State \Return $(z', \log(|\det(J)|))$
\EndProcedure
\end{algorithmic}
\caption{Automated Involutive MCMC}
\label{alg:auto-involutive-mcmc}
\end{algorithm*}

@@

At the most basic level, a probabilistic program is a program that makes random choices. 
Any such program induces a probability distribution over its possible \textit{execution traces}, records of each random choice it makes.
If each random choice is associated with a unique \textit{address} from the set of
dictionary keys $\mathcal{K}$, 
then these traces can be viewed as finite dictionaries, mapping the address of each random choice to its value. 
The distribution induced by a probabilistic program over
its execution traces can thus be understood as a measure on the space $(\mathcal{D}, \Sigma_\mathcal{D}, \mu_\mathcal{D})$ introduced in the previous section. Furthermore, densities of trace distributions with respect to $\mu_\mathcal{D}$ are typically easy to compute.

@@

%the set of addresses $\mathcal{K}$ is partitioned into two spaces $\mathcal{K}_1$ and $\mathcal{K}_2$\\

@@

%then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{k \in S} X_k \}$\\

@@

%\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{k \in S_1}\}
